Unshuffling Permutations

### 2015-09-17 article.tex
Suite modifs intro

Given three words $u$, $v$, and $w$, $u$ is said to a \emph{shuffle} of
$v$ and $w$ if it can be formed by interleaving the letters from $v$ and
$w$ in a way that maintains the left-to-right ordering of the letters
from each word. In addition to purely combinatorial questions, the
shuffle product of words leads to the following purely computational
questions:
\begin{enumerate}
    \item \label{item:problem_1}
    Given three words $u$, $v$, and $w$, decide if $u$ is a shuffle of
    $v$ and $w$.
    \item \label{item:problem_2}
    Given a word $u$, decide if there exist two words $v$ and $w$ such
    that $u$ is a shuffle of $v$ and $w$.
\end{enumerate}
Problem~\ref{item:problem_1} is in $\Pclass$ (it is a classical textbook
exercise to design an efficient dynamic programming algorithm that
determines whether $u$ is a shuffle of $v$ and $w$), while
Problem~\ref{item:problem_2} is in $\NPclass$ as shown
in~\cite{Rizzi:Vialette:CSR:2013}. Analogous questions can be stated to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle). A square
is a word $u$ with the property that there exists a word $v$ such that
$u$ is a shuffle of $v$ with itself.
\smallskip

This paper is intended to study a natural generalization of the shuffle
to permutations. Roughly speaking, given three permutations $\pi$, $\mu$,
and $\nu$, $\pi$ is said to be a shuffle of $\mu$ and $\nu$ if $\pi$
(viewed as a word) is a shuffle of two words that are order-isomorphic
to $\mu$ and $\nu$.
%For
%example, $\pi$ is the shuffle of $\mu=$ and $\nu=$ as $\pi$. A
%permutation $\pi$ is said to be a square (w.r.t. the shuffle) if there
%exists a permutation $\mu$ such that $\pi$ is a shuffle of $\mu$ with
%itself. For example, $\pi = 241356$ is a square as it is the shuffle of
%$213$ with itself (\emph{e.g.},
%$\pi = \dot{2}{4}\dot{1}{3}{5}\dot{6}$, where the $\dot{x}$'s
%denote the second occurrence), whereas $\pi = 4123$ is not.
Our intention in this paper is to to study this shuffle product of
permutations both from a combinatorial and from a computational point of
view by focusing on {\em square} permutations, that are permutations
$\pi$ being in the shuffle of a permutation $\mu$ with itself. The
shuffle product defined and considered in this paper is different from
other existing and well-known shuffle products on permutations. For
instance, in~\cite{DHT:IJAC:2002}, the authors define the
{\em shifted shuffle product}. In this product $\pi$ is a shuffle of
$\mu$ and $\nu$ if $\pi$ is the shuffle, as words, of $\mu$ and the
word obtained by incrementing all letters of $\nu$ by the size of $\mu$.
It is a simple exercise to prove that, given three permutations $\pi$,
$\mu$, and $\nu$, deciding if $\pi$ is a in the shifted shuffle of
$\mu$ and $\nu$ is in $\Pclass$.
\smallskip

Let us now  briefly review the key results for words that arise in our context.
%Formally, the shuffle of two words $u$ and $v$ of $A^*$ is the language
%$u \shuffle v$ consisting of all words $u_1 v_1 u_2 v_2 \ldots u_k v_k$,
%where $k \geq 0$ and the $u_i$ and the $v_i$ are the words  of $A^*$
%such that $u = u_1 u_2 \ldots u_k$ and $v = v_1 v_2 \ldots v_k$. We
%first briefly review the key results for words that arise in our context.
Given three words $u$, $v_1$ and $v_2$, it can be tested in
$O\left(|u|^2 / \log(|u|)\right)$ time whether or not
$u$ is in $v_1 \shuffle v_2$~\cite{Leeuwen:Nivat:IPL:1982}.
To the best of our knowledge, the first $O(|u|^2)$ time algorithm
appeared in~\cite{Mansfield:DAM:1983}. This algorithm can easily be
extended to check in polynomial-time whether or not a word is the
shuffle of any fixed number of given words. The shuffle $u \shuffle v$
of words $u$ and $v$ can be computed in
\begin{equation}
    O\left((|u|+|v|) \; \binom{|u|+|v|}{|u|}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}.
An improvement and generalization has been proposed in
\cite{Allauzen:IGM:2000}, where it is proved that, given words
$u_1, u_2, \ldots, u_k$, the shuffle $\shuffle_{i=1}^{k} u_i$ can be
computed in
\begin{equation}
    O\left(\binom{|u_1|+|u_2|+\ldots+|u_k|}{|u_1|,|u_2|,\ldots,|u_k|}\right)
\end{equation}
time. Given words $u, v_1, v_2, \ldots, v_n \in A^*$, it is however \NPC
to decide whether or not $u \in \shuffle_{i=1}^{k} v_i$.
\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}.
This remains true even if the alphabet has size $3$~\cite{Warmuth:Haussler:JCSS:1984}.
Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that this problem remains
\NPC even if all
words $v_1, v_2, \ldots, v_n$ are identical,
thereby proving that,
for two words $u$ and $v$, it is \NPC to decide whether or not $u$
is in the iterated shuffle of $v$.
Again, this remains true even if the alphabet has size $3$.
As for squares, given a word $z$ it is \NPC to determine whether
there exists a word $x$ such that $z$ is the shuffle of $x$ with itself
\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}.
\smallskip

This paper is organized as follows...

@@

There are a number of further directions of investigation in this
general subject. We mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, by
Proposition~\ref{prop:bijection_binary_to_permutations_squares},
how many binary strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?)
% Given two
% permutations $\pi$ and $\sigma$, how hard is the problem of deciding
% whether $\pi$ is the union of two patterns order-isomorphic to $\sigma$?
\smallskip

>>>>

Given three words $u$, $v$, and $w$, $u$ is said to a \emph{shuffle} of
$v$ and $w$ if it can be formed by interleaving the letters from $v$ and
$w$ in a way that maintains the left-to-right ordering of the letters
from each word. In addition to purely combinatorial questions, the
shuffle product of words leads to the following computational questions:
\begin{enumerate}
    \item \label{item:problem_1}
    Given three words $u$, $v$, and $w$, decide if $u$ is a shuffle of
    $v$ and $w$.
    \item \label{item:problem_2}
    Given a word $u$, decide if there exist two words $v$ and $w$ such
    that $u$ is a shuffle of $v$ and $w$.
\end{enumerate}
Problem~\ref{item:problem_1} is in $\Pclass$ (it is a classical
textbook exercise to design an efficient dynamic programming algorithm
that determines whether $u$ is a shuffle of $v$ and $w$), while
Problem~\ref{item:problem_2} is $\NPclass$-complete as shown
in~\cite{Rizzi:Vialette:CSR:2013}. Analogous questions can be stated to
decide if a word $u$ is a \emph{square} (w.r.t. the shuffle), that is
a word $u$ with the property that there exists a word $v$ such that
$u$ is a shuffle of $v$ with itself.
\smallskip

This paper is intended to study a natural generalization $\SHUFFLE$
of the shuffle
of permutations. Roughly speaking, given three permutations $\pi$, $\mu$,
and $\nu$, $\pi$ is said to be a {\em shuffle} of $\mu$ and $\nu$ if $\pi$
(viewed as a word) is a shuffle of two words that are order-isomorphic
to $\mu$ and $\nu$.
%For
%example, $\pi$ is the shuffle of $\mu=$ and $\nu=$ as $\pi$. A
%permutation $\pi$ is said to be a square (w.r.t. the shuffle) if there
%exists a permutation $\mu$ such that $\pi$ is a shuffle of $\mu$ with
%itself. For example, $\pi = 241356$ is a square as it is the shuffle of
%$213$ with itself (\emph{e.g.},
%$\pi = \dot{2}{4}\dot{1}{3}{5}\dot{6}$, where the $\dot{x}$'s
%denote the second occurrence), whereas $\pi = 4123$ is not.
Our intention in this paper is to study this shuffle product of
permutations both from a combinatorial and from a computational point of
view by focusing on {\em square} permutations, that are permutations
$\pi$ being in the shuffle of a permutation $\mu$ with itself. The
shuffle product defined and considered in this paper is different from
other existing and well-known shuffle products on permutations. For
instance, in~\cite{DHT:IJAC:2002}, the authors define the
{\em shifted shuffle product}. For this product, $\pi$ is a shuffle of
$\mu$ and $\nu$ if $\pi$ is in the shuffle, as words, of $\mu$ and the
word obtained by incrementing all the letters of $\nu$ by the size of $\mu$.
It is a simple exercise to prove that, given three permutations $\pi$,
$\mu$, and $\nu$, deciding if $\pi$ is a in the shifted shuffle of
$\mu$ and $\nu$ is in $\Pclass$.
\smallskip

Let us now  briefly review the key results for words that arise in our context.
%Formally, the shuffle of two words $u$ and $v$ of $A^*$ is the language
%$u \shuffle v$ consisting of all words $u_1 v_1 u_2 v_2 \ldots u_k v_k$,
%where $k \geq 0$ and the $u_i$ and the $v_i$ are the words  of $A^*$
%such that $u = u_1 u_2 \ldots u_k$ and $v = v_1 v_2 \ldots v_k$. We
%first briefly review the key results for words that arise in our context.
Given three words $u$, $v_1$, and $v_2$, it can be tested in
$O\left(|u|^2 / \log(|u|)\right)$ time whether or not
$u$ is in the shuffle $v_1 \shuffle v_2$ of $v_1$ and
$v_2$~\cite{Leeuwen:Nivat:IPL:1982}.
To the best of our knowledge, the first $O(|u|^2)$ time algorithm
appeared in~\cite{Mansfield:DAM:1983}. This algorithm can easily be
extended to check in polynomial-time whether or not a word is the
shuffle of any fixed number of given words. The shuffle $u \shuffle v$
of words $u$ and $v$ can be computed in
\begin{equation}
    O\left((|u|+|v|) \; \binom{|u|+|v|}{|u|}\right)
\end{equation}
time~\cite{Spehner:TCS:1986}.
An improvement and generalization has been proposed in
\cite{Allauzen:IGM:2000}, where it is proved that, given words
$u_1$, $u_2$, \dots, $u_k$, the shuffle $\shuffle_{i=1}^{k} u_i$ can be
computed in
\begin{equation}
    O\left(\binom{|u_1|+|u_2|+\ldots+|u_k|}{|u_1|,|u_2|,\ldots,|u_k|}\right)
\end{equation}
time. Given words $u$, $v_1$, $v_2$, \dots, $v_n$, it is however \NPC
to decide whether or not $u$ is in $\shuffle_{i=1}^{k} v_i$.
\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}.
This remains true even if the ground alphabet has size $3$~\cite{Warmuth:Haussler:JCSS:1984}.
Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that this problem remains
\NPC even if all
words $v_1, v_2, \ldots, v_n$ are identical,
thereby proving that,
for two words $u$ and $v$, it is \NPC to decide whether or not $u$
is in the iterated shuffle of $v$.
Again, this remains true even if the ground alphabet has size $3$.
As for squares, given a word $z$ it is \NPC to determine whether
there exists a word $x$ such that $z$ is the shuffle of $x$ with itself
\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}.
\smallskip

This paper is organized as follows. In
Section~\ref{section:Shuffle product on permutations} we provide a precise
definition of $\SHUFFLE$. This definition passes
through the preliminary definition of an operator $\Delta$ to unshuffle
permutations. This operator is in fact a coproduct, endowing the space
of all permutations with a coalgebra structure. By duality, the
unshuffling operator $\Delta$ leads to the definition of our shuffle operation on
permutations. This approach has many advantages. First, some combinatorial
properties of $\SHUFFLE$ depend on properties of $\Delta$
and are more easy to prove on the coproduct
side. Second, this way of doing allows to obtain a clear description of
the multiplicities of the elements appearing in the shuffle of two
permutations, which are worthy of interest.
Section~\ref{section:Binary square words and permutations}
is devoted to show that the problems related to the shuffle of words has
links with the shuffle of permutations. In particular, we show that
binary words that are square are in one-to-one bijection with square
permutations (Proposition~\ref{prop:bijection_binary_to_permutations_squares}).
Next, Section~\ref{section:Algebraic issues} presents some algebraic
and combinatorial properties of~$\SHUFFLE$. In particular, we show that
$\SHUFFLE$ is associative and commutative
(Proposition~\ref{prop:shuffle_associative_commutative}), and that
if a permutation if a square, its mirror, complement, and inverse also
are square (Proposition~\ref{prop:square_stability}). Finally,
Section~\ref{section:Algorithmic issues} presents
the most import result of this paper: the fact that deciding if
a permutation is a square is $\NPclass$-complete
(Proposition~\ref{proposition:hardness}).

@@

There are a number of further directions of investigation in this
general subject. We mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, by
Proposition~\ref{prop:bijection_binary_to_permutations_squares},
how many binary strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?
% Given two
% permutations $\pi$ and $\sigma$, how hard is the problem of deciding
% whether $\pi$ is the union of two patterns order-isomorphic to $\sigma$?
\smallskip
